
#Exercise 3

##Steps 1-2

###Background (Brief description of data)
The data is collected during the 2005-2006 school year from two public schools (secondary education) in Portugal.
The data include student grades, demographic, social and school related variables, which were collected by using school reports and questionnaires.Among these is also alcohol use. Two datasets are provided regarding the performance in two distinct subjects: Mathematics (mat) and Portuguese language (por). These two have been combined into one file.

Variables G1, G2, G3 are the grades from period 1 to 3 (3 is final grade). Range is from 0-20 and an description of the range of points is in table 1.

####Table 1.Grades in Portugal (and France)

| Excellent | Good | Satisfactory | Sufficient | Fail |
|-----------|------|--------------|------------|------|
|16-20p | 14-15p | 12-13p | 10-11p | 0-9p | fail |


```{r message=FALSE}
#Read the joined student alcohol consumption data into R
alc <- read.delim("~/GitHub/IODS-project/data/alc2018.txt")
dim(alc)
#str(alc)
#Observations: 382
#Variables: 35

#Print out the names of the variables
colnames(alc)

```

##Step 3-4

###Personal hypotheses

I chose the following interesting variables from the datset: "sex", "adress"" (student's home address type, rural or urban), "famrel"  (from 1 - very bad to 5 - excellent) and higher (wants to take higher education (binary: yes or no)). My personal hypotheses regarding these variables and their effect on alcohol consuption is:

Students  of male sex, with poor family relationships,that don't want to make a higher eduction and live in an urban environment drink more alcohol.

###Numerical and graphical exploration of the variables

Based on a numerical exploration of the data, it seems that familial relationships are slightly better among non-high alcohol users (table 2a), as hypothesized. However, the mean  alcohol intake seem to be higher among persons living in an rural envoronment(2b), which was unexpected. Those that wanted to take an higher eductation and persons of male sex had on average a higher alcohol intake (Table 2c-d). The same is also clear from the distribution figures of the selected variables (figure1a-d)

```{r echo=F, message=FALSE, warning=FALSE, fig.height=8, fig.width=8}

library(dplyr);library(ggplot2);library(tidyr);library(knitr);library(ggpubr)

#subsetting interesting variable for numerical data summarizes
my_var_names <- c("address", "famrel", "higher", "sex", "alc_use")
myalc_var <- subset(alc, select=my_var_names)


# produce summary statistics by group
kable(alc %>% group_by(high_use) %>% 
        summarise(count = n(),mean_famrel = mean(famrel)), 
      caption=" Table 2a. mean value of familial relationships among high vs low users of alcohol")

kable(alc %>% group_by(address) %>% 
        summarise(count = n(),mean_alc_use =mean(alc_use)), 
      caption = "Table 2b. mean value of alcohol use among persons with a rural or urban address")

kable(alc %>% group_by(higher) %>% summarise(count = n(),mean_alc_use =mean(alc_use)), 
      caption = "Table 2c. mean value of alcohol use divided by persons wanting to take higher eduction")

kable(alc %>% group_by(sex) %>% summarise(count = n(), mean_alc_use =mean(alc_use)),
      caption = "Table 2d. mean value of alcohol use divided by male and female")

                     
#check distribution of selected variables 
gather(myalc_var) %>% ggplot(aes(value), fill=high_use) + facet_wrap("key", scales = "free") +geom_histogram(stat = "count")+theme_minimal()


ad<- ggplot() + 
  geom_density(data=alc, aes(x=alc_use, group=address, fill=address),alpha=0.5, adjust=2) + 
  xlab("Alcohol intake (1-5)") +
  ylab("Density") +ggtitle("Alcohol intake divided by address\n(R=Rural, U=Urban) ")

sex<- ggplot() + 
  geom_density(data=alc, aes(x=alc_use, group=sex, fill=sex),alpha=0.5, adjust=2) + 
  xlab("Alcohol intake (1-5)") +
  ylab("Density") +ggtitle("Alcohol intake divided by gender\n(M=MAle, F=Female) ")

hig<- ggplot() + 
  geom_density(data=alc, aes(x=alc_use, group=sex, fill=sex),alpha=0.5, adjust=2) + 
  xlab("Alcohol intake (1-5)") +
  ylab("Density") +ggtitle("Alcohol intake divided by want to take\nhigher education")

famre<- ggplot() + 
  geom_density(data=alc, aes(x=famrel, group=high_use, fill=high_use),alpha=0.5, adjust=2) + 
  xlab("Familial relationships (1-5, poor-good)") +
  ylab("Density") +ggtitle("familial relationships divided by high alcohol intake\n(alc_use over 2)")

paste("Figure. 1")
ggarrange(ad, sex, hig, famre, labels=c("1a", "1b", "1c", "d"))


```

##step 5
###Logistic regression

Based on the summary of the fitted model, familial relationships are negatively associated with high alcohol use (P=0.02), having an urban address is negatively associatiod with alcohol but not on a statistically significant level (P=0.08), male sex is positively associated with high alcohol use (P=0.0002) and wanting to take a higher education is negatively associated (but not statistically significant!) with high alcohol use. The z- and p-values are based on the wald test, which is used in the model to test whether the pairwise difference between the coefficient and the reference class and the other class is different from zero or not (categorical values). 

Table 3 contains the Odds ratios of the modelled variables (and 95%CI), which were calculated by taking the exponent of the coefficients. By looking at these, we can say that

*For each increase(=improvement)in familial relationships there was 27% decrease in high use of alcohol, with the true population effect between 43% and 6%. 
*Persons having an urban address were 28% less likely to be high users of alcohol, however this observation was not statistically significant.
*Persons of male sex were 142% more likely to be high user of alkohol (true population effect between 52% and 285%).
*Persons who wanted to take a higher education were 44% less likely high alcohol users, however this observation was not statistically significant.


```{r warnings=F, message=F}
glm_res <- glm(high_use ~ famrel + address +sex+higher, data = alc, family = "binomial")
library(knitr)
#summary of the model
summary(glm_res)

#OR and 95%CIs
ORs_CI <-cbind(t(t(exp(coef(glm_res)))), exp(confint.default(glm_res)))
colnames(ORs_CI)[1] <- "OR"
kable(ORs_CI, caption = "Table 3. Odds Ratio and 95%CI")


```

##Step 6
###The predictive power of the model
A 2x2 cross tabulation of predictions versus the actual values is found in table 4 and  a graphic visualizing both the actual values and the predictions is in figure 4. The total proportion of inaccurately classified individuals (= the training error) is found in table 5.

The average number of wrong predictions in the training data was 0.28

```{r}

library(dplyr)
#fit a logistic regression with my sig variables
glm_res2 <- glm(high_use ~ famrel+sex, data = alc, family = "binomial")

#predict probability of high use
probabilities <- predict(glm_res2, type = "response")

# add the predicted probabilities to 'alc'
alc <- mutate(alc, probability = probabilities)

# use the probabilities to make a prediction of high_use
alc <- mutate(alc, prediction = probability > 0.5)

# see the last ten original classes, predicted probabilities, and class predictions
select(alc, famrel, sex, high_use, probability, prediction) %>% tail(10)

# tabulate the target variable versus the predictions
kable(table(high_use = alc$high_use, prediction = alc$prediction), caption = "Table 4. high alcohol use vs predictions")


# access dplyr and ggplot2
library(dplyr); library(ggplot2)

# initialize a plot of 'high_use' versus 'probability' in 'alc'
g <- ggplot(alc, aes(x = probability, y = high_use,col=prediction))

# define the geom as points and draw the plot
g +geom_point()+ggtitle("Figure 4. high use of alochol vs probability of high alcohol use")

# tabulate the target variable versus the predictions
kable(table(high_use = alc$high_use, prediction = alc$prediction), caption = "Table 5. high alcohol use vs predictions")

kable(table(high_use = alc$high_use, prediction = alc$prediction)  %>% prop.table() %>% addmargins(), caption = "Table 6. High alcohol use vs predictions")


# define a loss function (average prediction error)
loss_func <- function(class, prob) {
  n_wrong <- abs(class - prob) > 0.5
  mean(n_wrong)
}

# call loss_func to compute the average number of wrong predictions in the (training) data
loss_func(class = alc$high_use, prob = alc$probability)


```

##step 7
###10-fold Cross-validation
Bonus: Perform 10-fold cross-validation on your model. Does your model have better test set performance (smaller prediction error using 10-fold cross-validation) compared to the model introduced in DataCamp (which had about 0.26 error). Could you find such a model?

```{r}

```

##step 8
###Cross-validation ofthe performance of different logistic regression models 
Super-Bonus: Perform cross-validation to compare the performance of different logistic regression models (= different sets of predictors). Start with a very high number of predictors and explore the changes in the training and testing errors as you move to models with less predictors. Draw a graph displaying the trends of both training and testing errors by the number of predictors in the model. 

```{r}

```

?cbind(
